---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 150
)
```

# aapalette

<!-- badges: start -->
<!-- badges: end -->

**aapalette** provides a curated set of amino-acid colour palettes for protein
sequence figures in R. Every colour is loaded from a single bundled JSON source
of truth (`inst/extdata/aa_palettes.json`), so the R, Python (`pyaapalette`) and
Jalview (`jalaapalette`) siblings ship identical scheme IDs, hex values and
residue handling.

## Installation

```r
# install.packages("remotes")
remotes::install_github("aapalette/aapalette")
```

## The schemes

Three palettes are **new**, from the AApalette amino-acid colour alphabet
(this project):

| ID | For | Released |
|----|-----|----------|
| `hue` | normal vision | CC-BY-4.0 |
| `redgreen` | deuteranopia & protanopia (red-green CVD) safe | CC-BY-4.0 |
| `tritan` | tritanopia safe | CC-BY-4.0 |

Seven are **community-standard** schemes, reproduced and attributed:

| ID | Source |
|----|--------|
| `clustal` | Clustal X / Jalview |
| `zappo` | Zappo / Jalview |
| `taylor` | Taylor (1997) / Jalview |
| `lesk` | Lesk, *Introduction to Protein Architecture* |
| `cinema` | CINEMA (Parry-Smith *et al.* 1998) |
| `rasmol` | RasMol amino colour scheme |
| `shapely` | RasMol / Jmol shapely (Fletterick *Shapely* models) |

```{r schemes}
library(aapalette)
aa_schemes()
```

## Usage

Get a palette as a named character vector in canonical residue order
(`A C D E F G H I K L M N P Q R S T V W Y`):

```{r palette}
aa_palette("hue")
```

Optionally append the documented defaults for unknown/ambiguous codes and gaps:

```{r palette-defaults}
aa_palette("clustal", include_unknown = TRUE, include_gap = TRUE)
```

Scheme metadata, including the CIEDE2000 ΔE minima for the CVD-aware palettes:

```{r info}
aa_scheme_info("redgreen")$min_deltaE
```

Look up colours for an arbitrary sequence (lower-case, unknown codes and gaps
are handled automatically):

```{r colour}
aa_colour(strsplit("ACDEKR-X", "")[[1]], "hue")
```

### Swatches

```{r swatch-all, fig.height = 12, fig.width = 9}
op <- par(mfrow = c(10, 1))
for (id in aa_schemes()$id) plot_aa_palette(id)
par(op)
```

### ggplot2

`scale_colour_aa()` / `scale_color_aa()` and `scale_fill_aa()` are discrete
manual scales keyed on residue letters. Unknown residues and gaps fall back to
the documented defaults.

```{r ggplot, fig.height = 2, fig.width = 7}
library(ggplot2)
df <- data.frame(residue = factor(names(aa_palette("hue"))), x = 1:20, y = 1)
ggplot(df, aes(x, y, fill = residue, label = residue)) +
  geom_tile() +
  geom_text() +
  scale_fill_aa("hue") +
  theme_void() +
  theme(legend.position = "none")
```

## Recommended schemes

```{r recommended}
aa_recommended()
```

Use **`hue`** for normal vision, **`redgreen`** for red-green colour-vision
deficiency, and **`tritan`** for tritanopia.

## Colour-vision caveat

> No 20-colour palette is safe for all deficiencies at once; for robust figures
> pair colour with the residue letter (redundant coding). ΔE values are
> CIEDE2000 minima.

The swatch helper labels every residue precisely to encourage this redundant
coding.

## Deliberate exclusions

Polychrome, Green-Armytage, and Biotite/Gecos (flower/blossom/sunset) palettes,
and any personal/agent palettes, are intentionally **not** included.

## Licences & attribution

- **Code:** MIT (see `LICENSE` / `LICENSE.md`).
- **Colour data:** the palettes in `inst/extdata/aa_palettes.json` — including
  the new `hue`, `redgreen` and `tritan` palettes — are released under
  **CC-BY-4.0** (see `inst/extdata/LICENSE-CC-BY-4.0.txt`).

If you use the new palettes, please cite the AApalette amino-acid colour
alphabet:

> AApalette: a property-coherent, colour-vision-aware amino-acid colour
> alphabet. *Methods write-up forthcoming* (placeholder citation), 2026.

The classical schemes are attributed to their original sources as listed in the
table above and in each scheme's `source` field.

## Data provenance

The bundled `inst/extdata/aa_palettes.json` is the single source of truth and is
copied verbatim from the project-root `aa_palettes.json`. Never edit hex values
by hand; see `data-raw/regenerate.R` to refresh the bundled copy. The test suite
asserts that the package's palettes equal the JSON exactly.
